Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-28 ms/spaceness.tex
peter edited mostly through

@@

The confounding effects of geographic structure are well-known, 
and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a 5\% FDR significance threshold. 
However, a substantial number of SNPs -- up to 1.5\% of SNPs -- still surpass this threshold (and thus would be false positives in a GWAS),
especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
\caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-log10(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
%A: example environmental distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: Proportion of significant SNPs after FDR correction for GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect deflated $p$ values.
    \plr{suggestion: (a) log-scale the y axes in (C) and add another y-axis on the right showing (rough number of human genome SNPs above 5\% times this percentage) so we can see absolute numbers}
}
\label{fig:gwas}
\end{figure*}

\section{Discussion}
Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short. While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g., \cite{Wright1943,Rousset1997,Ringbauer2017,Barton2010}), the theoretical work is challenging and often untenable for realistic biological models. Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation. In particular we examine three axes of variation across three different sampling strategies -- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS). We are in particular interested in asking how our empirical inferences from data might be affected by spatial processes. As we show below the answers seems to be - often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.

\subsection{Effects of Dispersal}
Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, and introduces subtle biases in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}). At the extreme low end of dispersal distance this can result in an up to three-fold increase in genetic diversity relative to random-mating expectations. These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g., inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling). Under random sampling the general pattern is similar to expectations of the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.

The patterns observed in sequence data reflect the effects of space on the underlying genealogy. Nearby individuals coalesce rapidly under limited dispersal and so are connected by short branch lengths, while distant individuals take much longer to coalesce than they would under random mating. Mutation and recombination events in our simulation both occur at a constant rate along branches of the genealogy, so the genetic distance and number of recombination events separating two individuals is simply a noisy estimate of the branch lengths connecting them. These genealogical patterns also relate directly to the site frequency spectrum. In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating. Tip branches (i.e., branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long. These patterns then create the biases we observed in the SFS -- the lowest frequency bins are deflated by the short branch lengths connecting nearby individuals, while mid-frequency bins are inflated by the long branches connecting local groups.

The genealogical patterns introduced by limited dispersal are particularly apparent in the distribution of haplotype block lengths (Figure \ref{fig:sumstats}). This is because identical-by-state tract lengths reflect the impacts of two processes acting along the branches of the underlying genealogy -- both mutation and recombination -- rather than just mutation as is the case when looking at the site frequency spectrum or related summaries. This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{ST}$ or $D_{xy}$ and space \citep{Rousset1997}. Indeed this basic logic underlies two recent studies explicitly estimating dispersal from the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}. Conversely, because haplotype-based measures of demography are particularly sensitive to variation in the underlying genealogy, inference approaches that assume random mating when analyzing the distribution of shared haplotype block lengths are likely to be strongly affected by spatial processes.

\subsection{Effects of Sampling}
One of the most important differences between random mating and spatial models is the effect of sampling: in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. As expected we found that random sampling provides the most accurate summary of genetic diversity across the landscape. However this strategy is often impractical for empirical studies. In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the ``point'' sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape. For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.

When sampling is clustered at points on a landscape and dispersal is limited, the sampled individuals will be more closely related than a random set of individuals. Average coalescence times of individuals collected at a locality will then be more recent and branch lengths shorter than expected by analyses assuming random mating. This leads to fewer mutations and recombination events occurring since their last common ancestor, causing a random set of individuals to share longer average IBS tracts and have fewer nucleotide differences. For some data summaries, such as Tajima's $D$, Watterson's $\Theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}). Inferring underlying demographic parameters from these summary statistics -- for example, estimating dispersal distance as the slope of a regression of $F_{ST}$ against the logarithm of spatial distance \citep{Rousset1997} -- may then be subject to bias if sampling is not random across the landscape.

However, the largest sampling effects we observed occurred in our ``midpoint'' sampling strategy. This model is meant to reflect a bias in sampling effort towards the middle of a species' range. In empirical studies this sampling strategy could arise if, for example, researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season. Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample. This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation that have often been ignored by analytic approaches focusing on infinite toroidal landscapes \citep{Felsenstein1975}. If individuals move randomly in a finite two-dimensional landscape then regions in the middle of the landscape receive migrants from all directions while those on the edge receive no migrants from at least one direction. The average number of new mutations moving into the middle of the landscape is then higher than the number moving into regions near the range edge, leading to higher heterozygosity and lower inbreeding coefficients ($F_{IS}$) away from range edges. Though here we used only a single parameterization of fitness decline at range edges we believe this is a general property of non-infinite landscapes as it has also been observed in previous studies simulating under lattice models \citep{Neel2013,Shirk2014}. For empirical studies this suggests that sampling only the middle of the landscape will produce a biased view of many aspects of the genetic variation in a population.

In summary, empirical researchers should collect individuals in as random a manner as practical and should not bias their sample towards the middle of a species' range. When sampling is clustered, summary statistics based on segregating sites (e.g., Watterson's $\Theta$ and Tajima's $D$), heterozygosity, or the distribution of long haplotype blocks, can be expected to depart significantly from what would be observed under random sampling. Comparing the results of analyses conducted on all individuals versus those limited to single individuals per locality may help to reveal any effect of sampling bias if sufficient samples are available. Alternatively, sampling strategy could be incorporated directly into a simulation-based inferential framework by, for example, setting the  distribution of samples in a simulation to mimic that used for the empirical target case. This could be readily achieved in a supervised machine learning or approximate Bayesian setting (CITES).

\subsection{Demography}
Classical population genetic models collapse many elements of life history variation into a single parameter, $N_{e}$, which is then taken to reflect the degree of variation present in the population when modeling the effects of selection or migration. Inferring $N_{e}$ in the past is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation when searching for signals of selection. Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20. However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}). All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.

@@

\subsection{GWAS}
Over the last twenty years genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species. This technique is increasingly applied to questions of human health through methods like polygenic risk scores that sum the effect sizes estimated from GWAS to predict an individual's phenotype or disease risk \citep{Khera2018}. The most common approach to GWAS, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype. As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.

Stepping back from the mechanics of GWAS specifically, the approach of quantitative genetics is to decompose the variance in phenotypes into environmental and genetic effects, e.g., 
\begin{equation}
P=G+E
\end{equation}
\begin{equation}
var(P)=var(G)+var(E)+cov(G,E)
\end{equation}
For GWAS in structured populations, the bias identified in many previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015} in which test statistics reflect population structure rather than phenotype association arises because of the presence of a positive covariance between genotypic and environmental variation (the third term above). When the environment and genotype covary across space their effects are confounded. Note this does not require interaction between genotype and environment (so-called GxE effects), which will introduce an additional source of bias. Here we refer simply to the fact that the allele frequency at a site will often covary with the environment when both breeding structure and the environment vary over space. To some degree the success of quantitative genetics in fields like agriculture is likely due to the absence of this covariation -- crop and animal breeding operations can be conducted so that the environment is identical (or nearly so) or randomized across populations. However in natural populations this situation is extremely unlikely. Most populations are structured by a combination of limited dispersal and geographic barriers, and nearly all environments vary over space. GWAS in natural populations is forced to confront the confounding effects of population structure and the environment directly.

Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation. However while this approach is quite useful, it does not truly separate the confounding signals of environment and spatially varying genotypes. In essence with a PC-corrected GWAS we are asking ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' In our simulations we observed that this procedure can fail under a variety of circumstances. If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association. Conversely when dispersal is high we found that PCA loses power to describe population structure before the spatial scale of dispersal breaks down the relationship between genotype and the environment. These effects were observed in all spatially varying environmental distributions, but were particularly pronounced for concentrated environmental effects in one region, as was also found in \cite{Mathieson2012}. As a result we can expect to see several thousand weak false-positive associations in a PC-corrected GWAS conducted on a human-sized genome in species with neighborhood sizes up to at least 1000.

This does not mean that GWAS is not useful, but does put some limits on the extent of valid interpretation. Very few of the associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff, suggesting that most of the very strong signals of association signals observed in studies of mono- or oligogenic traits are robust to stratification bias. Further, the most dramatic effects of stratification inflation we observed occurred at neighborhood sizes below 100 -- smaller than the vast majority of modern human populations (but see below for further discussion of empirical cases). However, as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations does leave residual signal of population structure in large-scale GWAS of polygenic traits. Indeed, studies in strongly structured species like \textit{Arabadopsis} have long relied on more sophisticated mixed model approaches to correcting for population structure for precisely this reason \citep{Aranzana2005,Sasaki2015}.

A second point that has received less attention in the literature is the issue of overcorrection in GWAS. If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of effect sizes. Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution. This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}. Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}\ak{i don't understand what the authors mean here -- how can we have a negative variance?}. Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.

In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled environments. One currently popular approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}. We suggest a straightforward alternative for species in which the primary axes of population differentiation is space (note this is likely not the case for many modern human populations): run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, the structure corrections are insufficient. This is essentially the approach taken in our ``clinal'' model (though we also include normally distributed variation in our phenotypes). Of course it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.  \cjb{what do you think of the second half here?}

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation? In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation. Though this sample is almost certainly biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density), we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference. These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation. Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in NE Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

>>>>

\plr{Moved from disco: integrate citations.}
Patterns of genetic variation are influenced by the fact that organisms are more likely, on average,
to live nearby to both their parents, in particular when dispersal and mate choice distances are short. 
Analytical work describing the genetics of populations distributed across continuous geography
(e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation})
the theoretical work is challenging and often untenable for realistic biological models. 
Also discuss lots of work on grids, e.g., \citet{Shirk2014,rousset2011likelihoodbased}.

@@

The confounding effects of geographic structure are well-known, 
and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a 5\% FDR significance threshold. 
However, a substantial number of SNPs -- up to 1.5\% of SNPs -- still surpass this threshold (and thus would be false positives in a GWAS),
especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction.

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
%A: example environmental distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: Proportion of significant SNPs after FDR correction for GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect deflated $p$ values.
    \plr{suggestion: (a) log-scale the y axes in (C) and add another y-axis on the right showing (rough number of human genome SNPs above 5\% times this percentage) so we can see absolute numbers}
}
\label{fig:gwas}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

In this study,
we have used efficient forward time population genetic simulations 
to describe the myriad influence of continuous geography on genetic variation. 
In particular, we examine how three main types of downstream empirical inference are affected by unmodeled spatial population structure
-- 1) population genetic summary statistics, 2) inference of population size history, and 3) genome-wide association studies (GWAS). 
As discussed above, 
space often matters (and sometimes dramatically),
both because of how samples are arranged in space, and because of the inherent patterns of relatedness established by geography.

%%%%%%%%%%
\subsection{Effects of Dispersal}

Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, 
and introduces strong distortions in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}). 
At the lowest dispersal distances, this can increase genetic diversity threefold relative to random-mating expectations. 
These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g., inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling). Under random sampling the general pattern is similar to expectations of the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.

The patterns observed in sequence data reflect the effects of space on the underlying genealogy. Nearby individuals coalesce rapidly under limited dispersal and so are connected by short branch lengths, while distant individuals take much longer to coalesce than they would under random mating. 
Mutation and recombination events in our simulation both occur at a constant rate along branches of the genealogy, 
so the genetic distance and number of recombination events separating sampled individuals simply gives a noisy picture of the genealogies connecting them. 
% These genealogical patterns also relate directly to the site frequency spectrum. 
% In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating. 
Tip branches (i.e., branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long. These patterns then create the biases we observed in the SFS -- the lowest frequency bins are deflated by the short branch lengths connecting nearby individuals, while mid-frequency bins are inflated by the long branches connecting local groups.

The genealogical patterns introduced by limited dispersal are particularly apparent in the distribution of haplotype block lengths (Figure \ref{fig:sumstats}). This is because identical-by-state tract lengths reflect the impacts of two processes acting along the branches of the underlying genealogy -- both mutation and recombination -- rather than just mutation as is the case when looking at the site frequency spectrum or related summaries. This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{ST}$ or $D_{xy}$ and geographic distance \citep{Rousset1997}. Indeed this basic logic underlies two recent studies explicitly estimating dispersal from the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}. Conversely, because haplotype-based measures of demography are particularly sensitive to variation in the underlying genealogy, inference approaches that assume random mating when analyzing the distribution of shared haplotype block lengths are likely to be strongly affected by spatial processes.

\subsection{Effects of Sampling}
One of the most important differences between random mating and spatial models is the effect of sampling: 
in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, 
but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. 
In most species,
the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site 
means that most studies follow something closest to the ``point'' sampling strategy we simulated, 
in which multiple individuals are sampled from nearby points on the landscape. 
For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. 
In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be extensively sampled.

When sampling is clustered at points on a landscape and dispersal is limited, the sampled individuals will be more closely related than a random set of individuals. Average coalescence times of individuals collected at a locality will then be more recent and branch lengths shorter than expected by analyses assuming random mating. This leads to fewer mutations and recombination events occurring since their last common ancestor, causing a random set of individuals to share longer average IBS tracts and have fewer nucleotide differences. For some data summaries, such as Tajima's $D$, Watterson's $\Theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}). 
Inferring underlying demographic parameters from these summary statistics 
-- unless the nature of the sampling is somehow taken into account 
-- will be subject to bias if sampling is not random across the landscape.

However, the largest sampling effects we observed occurred in our ``midpoint'' sampling strategy. 
This model is meant to reflect a bias in sampling effort towards the middle of a species' range. 
In empirical studies this sampling strategy could arise if, for example, 
researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season. 
Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample. 
This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
If individuals move randomly in a finite two-dimensional landscape then regions in the middle of the landscape receive migrants from all directions while those on the edge receive no migrants from at least one direction. The average number of new mutations moving into the middle of the landscape is then higher than the number moving into regions near the range edge, leading to higher heterozygosity and lower inbreeding coefficients ($F_{IS}$) away from range edges. Though here we used only a single parameterization of fitness decline at range edges we believe this is a general property of non-infinite landscapes as it has also been observed in previous studies simulating under lattice models \citep{Neel2013,Shirk2014}.

In summary, empirical researchers should collect individuals from across as much of the species' range as practical,
choosing samples at a variety of spatial scales.
Many summary statistics are designed for well-mixed populations,
and so provide different insights into genetic variation
when applied to different subsets of the population.
Applied to a cluster of samples, summary statistics based on segregating sites 
(e.g., Watterson's $\Theta$ and Tajima's $D$), 
heterozygosity, or the distribution of long haplotype blocks, 
can be expected to depart significantly from what would be obtained from a wider distribution of samples.
Comparing the results of analyses conducted on all individuals versus those limited to single individuals per locality 
can provide an informative contrast.

%%%%%%%%%%%
\subsection{Demography}

Classical results in population genetics collapse many aspects of life history into a single compound parameter, effective population size ($N_{e}$).
Inferring effective population size back through time is now a common goal of population genomic analyses 
and an important step in establishing baseline expectations of genetic variation, e.g., when searching for signals of selection. 
Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20. However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}). All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.

@@

%%%%%%%%%%%
\subsection{GWAS}

Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, 
both in humans and other species. 
This technique is increasingly applied to questions of human health through methods like polygenic risk scores 
that sum the effect sizes estimated from GWAS to predict an individual's phenotype or disease risk \citep{Khera2018}. 
The most common method, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, 
taking the slope of this regression as an estimate of the effect of the allele on the phenotype. 
As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.

Stepping back from specific methods,
the general goal is to identify alleles that are more correlated with the phenotype of interest
than one would expect due to chance.
Both phenotype and genotype are often spatially autocorrelated (for different reasons),
which makes spurious correlations a major obstacle,
both by creating false positives and reducing power to identify truly causal alleles 
\citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}.
It may be that the success of quantitative genetics in agriculture is in part because 
crop and animal breeding operations can be conducted so that the environment is identical (or nearly so) or randomized across populations. 
However, most natural populations are structured by a combination of limited dispersal and geographic barriers, 
and nearly all environments vary over space.

Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation. 
% This approach is extremely useful, but does not truly separate the confounding signals of environment and spatially varying genotypes. 
In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
In our simulations, we observed that this procedure can fail under a variety of circumstances. 
If dispersal is limited and environmental variation is clustered in space 
(i.e., corner or patchy distributions in our simulations), 
PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association. 
Conversely, as dispersal is increased, PCA loses power to describe population structure 
before spatial mixing breaks down the relationship between genotype and the environment. 
These effects were observed with all spatially correlated environmental patterns,
but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}. 
\plr{need to highlight a statement like this more}
As a result we can expect to see several thousand weak false-positive associations in a PC-corrected GWAS conducted on a human-sized genome in species with neighborhood sizes up to at least 1000.

This suggests some caveats for interpretation of GWAS results.
Very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff, 
suggesting that most of the very strong signals of association observed in studies of mono- or oligogenic traits are robust to these concerns.
Further, the most dramatic $p$-value inflations occurred at neighborhood sizes below 100 
-- smaller than most recent human populations (but see below for further discussion of empirical cases). 
However, as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, 
PC regression GWAS in modern human populations does leave residual signal of population structure in large-scale GWAS of polygenic traits. 
\plr{Is there any indication that mixed models do better about this? (Where's the ``for precisely this reason'' come from?)}
Indeed, studies in strongly structured species like \textit{Arabadopsis} have long relied on more sophisticated mixed model approaches to correcting for population structure for precisely this reason \citep{Aranzana2005,Sasaki2015}.

A second point that has received less attention in the literature is the issue of overfitting in GWAS. 
If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size. Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution. This is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}. Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
\ak{i don't understand what the authors mean here -- how can we have a negative variance?}
\plr{peter revisit this}
% Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.

In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, 
and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. 
Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}. 
We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
(note this is likely not the case for many modern human populations): 
run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. 
If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, 
the method is sensitive to false positives induced by spatial structure. 
This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). 
Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.  
\cjb{what do you think of the second half here?}
\plr{love it}

@@

%%%%%%%%%%%%
\subsection{Where are natural populations on this spectrum?}

How strong is the spatial patterning of genetic variation in extant species?
% For how much of the tree of life do spatial patterns circumscribe genomic variation? 
To get an idea of this, we gathered estimates of neighborhood size from a range of organisms (Table~\ref{table:NStable}).
Although this sample is almost certainly biased towards small-neighborhood species 
(because few studies have quantified neighborhood size in species with very high dispersal or population density), 
we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. 
At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. 
Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference. 
These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, 
for whom the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation. 
Many more species occur in a middle range of neighborhood sizes between 100 and 1000 
-- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. 
Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. 
If these species' ranges and population densities have been constant for a long period of time (unlike humans),
demographic inference methods that assume well-mixed populations may show minimal bias from spatial effects.
% While that is so, we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

%%%%%%%%%%%
\subsection{Future Directions and Limitations}
